Arriaga meets Kitagawa. life expectancy decompositions including population subgroups

### 2024-01-24 manuscript/manuscript.tex
trying out Rmd rticle for this one

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-apa      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: "Arriaga meets Kitagawa. life expectancy decompositions including population subgroups"

titlerunning: Arriaga meets Kitagawa

authors: 
  - prefix: Dr.
    firstname: Timothy
    #particle: van 
    lastname: Riffe
    #suffix: III
    #naturename: Poet Laureate
    #degrees: MSc, PhD
    email: tim.riffe@ehu.eus
    affiliation: [1,2,3]
    corresponding: TRUE
#    equalcont: These authors contributed equally to this work.

  # - firstname: Second
  #   lastname: Author
  #   affiliation: 2
  #   equalcont: These authors contributed equally to this work.

affiliations:
  - number: 1
    corresponding: TRUE
    info:
      orgdiv: Department of Sociology and Social Work
      orgname: University of the Basque Country (UPV/EHU)
    address:
        street: Barrio Sarriena s/n
        city: Leioa
        postcode: 48940
        state: Bizkaia
        country: Spain
  - number: 2
    corresponding: FALSE
    info:
      orgname: Ikerbasque (Basque Foundation for Science)
  - number: 3
    corresponding: FALSE
    info:
      orgdiv: Laboratory of Population Health
      orgname: Max Planck Institute for Demographic Research

keywords:
  - Decomposition
  - Mortality
  - Cause of death
  - Population Structure
  - Mortality Inequalities

# pacs: 
#   jel:
#     - "D8"
#     - "H51"
#   msc:
#     - "35A01"
#     - "65L10"

# Sample for unstructured abstract 
# abstract: |
#   The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.

# Sample for structured abstract
abstract: |
 **Background**: An Arriaga (1984) decomposition allows us to decompose differences in life expectancy into contributions due to mortality rate differences in each age. A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into effects from differences in structure and differences in the weighted value.

 **Objective:** Sometimes we would like to decompose a difference between two populations that are each composed of like-defined subpopulations. Said decomposition would produce effects for each age of each subpopulation, as well as a marginal effect of composition differences. I propose a straightforward analytic method to do this.

 **Methods:** In short, within-subpopulation life expectancy differences can be handled by the Arriaga method. The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method. The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation.

 **Results:** I show the proposed analytic method to be equivalent to a Horiuchi et al (2008) reframing of the same problem. Some mentionable properties: (i) There is no limit to the number of subpopulations, (ii) it is straightforward to incorporate cause-of-death information, (iii) composition is here only considered in the radix age. I currently have results for simulated mortality rates, but I promise to wrangle up an empirical application to demonstrate the method.

 **Conclusions:** The analytic decomposition method I propose is advantageous compared to a Horiuchi method for this problem purely for reasons of computational efficiency. This method could help further disentangle the effects of mortality and composition differences in explaining or clarifying paradoxes or secular change. I promise to think further about subpopulation weighting that might occur over all ages (using prevalence information).}

bibliography: bibliography.bib

header-includes: |
  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

@@

output: rticles::springer_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.extra = "")  # This forces knitr to label all figures.
```

# Introduction

An @arriaga1984measuring decomposition allows us to decompose changes in life expectancy into contributions due to mortality changes in different ages. The method was designed to be practical, and framed in terms of age lifetable columns expressed in discrete age groups. A well-known property of the method is that mortality changes in different ages need not be proportional. Derived contributions also sum exactly to the observed life expectancy difference. A not-well-known property of the method is that it is asymmetrical, in the sense that the absolute values of age-specific contributions depend on whether we compare population 1 with 2, or population 2 with 1. A further property of the method is that it is designed to work with homogeneous populations, meaning that populations 1 and 2 are each homogeneous, in the sense that each is expressed with only one lifetable.

A @kitagawa1955components decomposition allows us to decompose differences in a weighted mean due to differences in weights (structure) and differences in the value being weighted (often rates, but in our case life expectancies). This widely-used decomposition method is well-known to be exact in that the resulting structure and value components sum exactly to the observed difference in weighted means. The individual elements (ages, or life expectancies for us) of the value being weighted have identifiable effects. It is not well-known that the individual elements of the structure component do not have identifiable effects. Rather, the structure effects should be summed to a a marginal effect due to differences in structure.

Sometimes we have a situation where we would like to decompose a difference between two populations that are each composed of like-defined subpopulations. For example, life expectancy in France versus Spain, each with education-specific subpopulations. In such situations, a decomposition should tell us the contribution to the difference in overall life expectancy due to rate differences in each age in each subpopulation, and also separate an effect due to overall compositional change. In this paper, I would like to propose a straightforward analytic method to decompose in this way. In short, within-subpopulation life expectancy differences can be handled by the Arriaga method. The case of weighting together the life expectancies of subpopulations by way of a composed radix can be handled using the well-known Kitagawa method. The elements of the value component of the Kitagawa method tell us how to rescale the Arriaga results specific to each subpopulation so as to isolate the age-subpopulation-specific effects on the overall life expectancy difference. I will justify this rescaling, and show that the results of this procedure are fully consistent with the results of a @horiuchi2008decomposition reframing of the problem.

# Method

## Notation

Arriaga decompositions can be implemented in a number of relevant ways, per Riffe et. al. (2024), from which we here consider only symmetrical Arriaga decompositions. This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results. For this, we use the following lifetable columns:

```{=tex}
\begin{itemize}

@@

\end{itemize}
```
We also use the superscript $s$ to index subpopulations comprising the total population, and the superscript $t$ to index time points, or some other way of differentiating the total population. The index $e^{s,t}_0$ reads as "life expectancy at birth for subpopulation $s$ at time $t$", e.g. $e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group.

Then one can write Arriaga's standard decomposition method like as per eq~\eqref{eq_arriaga1}. To keep things simple, we assume a lifetable radix of 1, meaning $\ell_0 = 1$.

$$
\begin{aligned}
{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
 &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
\end{aligned}
$$

@@

# Declarations {.unnumbered}

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading \`Declarations':

-   Funding
-   Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
-   Ethics approval
-   Consent to participate
-   Consent for publication
-   Availability of data and materials
-   Code availability
-   Authors' contributions

\noindent If any of the sections are not relevant to your manuscript, please include the heading and write \`Not applicable' for that section.

::: {.flushleft latex="true"}
Editorial Policies for:

\noindent Springer journals and proceedings: <https://www.springer.com/gp/editorial-policies>

\noindent Nature Portfolio journals: <https://www.nature.com/nature-research/editorial-policies>

\noindent \textit{Scientific Reports}: <https://www.nature.com/srep/journal-policies/editorial-policies>

\noindent BMC journals: <https://www.biomedcentral.com/getpublished/editorial-policies>
:::

::: {.appendices latex="true"}
# Section title of first appendix {#secA1}

@@

For submissions to Nature Portfolio Journals please use the heading "Extended Data".
:::

>>>>

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-apa,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% Per the spinger doc, new theorem styles can be included using built in style, 
%% but it seems the don't work so commented below
%\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

@@

\raggedbottom

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\title[Arriaga meets Kitagawa]{Arriaga meets Kitagawa. life expectancy
decompositions including population subgroups}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1,2,3]{\pfx{Dr.} \fnm{Timothy} \sur{Riffe} }\email{\href{mailto:tim.riffe@ehu.eus}{\nolinkurl{tim.riffe@ehu.eus}}}

  \affil*[1]{\orgdiv{Department of Sociology and Social
Work}, \orgname{University of the Basque Country
(UPV/EHU)}, \orgaddress{\city{Leioa}, \country{Spain}, \postcode{48940}, \state{Bizkaia}, \street{Barrio
Sarriena s/n}}}
  \affil[2]{\orgname{Ikerbasque (Basque Foundation for Science)}}
  \affil[3]{\orgdiv{Laboratory of Population Health}, \orgname{Max
Planck Institute for Demographic Research}}

\abstract{\textbf{Background}: An Arriaga (1984) decomposition allows us
to decompose differences in life expectancy into contributions due to
mortality rate differences in each age. A Kitawaga (1955) decomposition
allows us to decompose differences in a weighted mean into effects from
differences in structure and differences in the weighted value.

\textbf{Objective:} Sometimes we would like to decompose a difference
between two populations that are each composed of like-defined
subpopulations. Said decomposition would produce effects for each age of
each subpopulation, as well as a marginal effect of composition
differences. I propose a straightforward analytic method to do this.

\textbf{Methods:} In short, within-subpopulation life expectancy
differences can be handled by the Arriaga method. The case of weighting
together the life expectancies of subpopulations by way of a composed
radix can be handled using the well-known Kitagawa method. The elements
of the value component of the Kitagawa method tell us how to rescale the
Arriaga results specific to each subpopulation.

\textbf{Results:} I show the proposed analytic method to be equivalent
to a Horiuchi et al (2008) reframing of the same problem. Some
mentionable properties: (i) There is no limit to the number of
subpopulations, (ii) it is straightforward to incorporate cause-of-death
information, (iii) composition is here only considered in the radix age.
I currently have results for simulated mortality rates, but I promise to
wrangle up an empirical application to demonstrate the method.

\textbf{Conclusions:} The analytic decomposition method I propose is
advantageous compared to a Horiuchi method for this problem purely for
reasons of computational efficiency. This method could help further
disentangle the effects of mortality and composition differences in
explaining or clarifying paradoxes or secular change. I promise to think
further about subpopulation weighting that might occur over all ages
(using prevalence information).\}}

\keywords{Decomposition, Mortality, Cause of death, Population
Structure, Mortality Inequalities}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

An \citet{arriaga1984measuring} decomposition allows us to decompose
changes in life expectancy into contributions due to mortality changes
in different ages. The method was designed to be practical, and framed
in terms of age lifetable columns expressed in discrete age groups. A
well-known property of the method is that mortality changes in different
ages need not be proportional. Derived contributions also sum exactly to
the observed life expectancy difference. A not-well-known property of
the method is that it is asymmetrical, in the sense that the absolute
values of age-specific contributions depend on whether we compare
population 1 with 2, or population 2 with 1. A further property of the
method is that it is designed to work with homogeneous populations,
meaning that populations 1 and 2 are each homogeneous, in the sense that
each is expressed with only one lifetable.

A \citet{kitagawa1955components} decomposition allows us to decompose
differences in a weighted mean due to differences in weights (structure)
and differences in the value being weighted (often rates, but in our
case life expectancies). This widely-used decomposition method is
well-known to be exact in that the resulting structure and value
components sum exactly to the observed difference in weighted means. The
individual elements (ages, or life expectancies for us) of the value
being weighted have identifiable effects. It is not well-known that the
individual elements of the structure component do not have identifiable
effects. Rather, the structure effects should be summed to a a marginal
effect due to differences in structure.

Sometimes we have a situation where we would like to decompose a
difference between two populations that are each composed of
like-defined subpopulations. For example, life expectancy in France
versus Spain, each with education-specific subpopulations. In such
situations, a decomposition should tell us the contribution to the
difference in overall life expectancy due to rate differences in each
age in each subpopulation, and also separate an effect due to overall
compositional change. In this paper, I would like to propose a
straightforward analytic method to decompose in this way. In short,
within-subpopulation life expectancy differences can be handled by the
Arriaga method. The case of weighting together the life expectancies of
subpopulations by way of a composed radix can be handled using the
well-known Kitagawa method. The elements of the value component of the
Kitagawa method tell us how to rescale the Arriaga results specific to
each subpopulation so as to isolate the age-subpopulation-specific
effects on the overall life expectancy difference. I will justify this
rescaling, and show that the results of this procedure are fully
consistent with the results of a \citet{horiuchi2008decomposition}
reframing of the problem.

\hypertarget{method}{%
\section{Method}\label{method}}

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

Arriaga decompositions can be implemented in a number of relevant ways,
per Riffe et. al.~(2024), from which we here consider only symmetrical
Arriaga decompositions. This implies following the original formulas
twice, once in each direction, and averaging the sign-adjusted results.
For this, we use the following lifetable columns:

\begin{itemize}

@@

\end{itemize}

We also use the superscript \(s\) to index subpopulations comprising the
total population, and the superscript \(t\) to index time points, or
some other way of differentiating the total population. The index
\(e^{s,t}_0\) reads as ``life expectancy at birth for subpopulation
\(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
expectancy in 1990 of a low education group.

Then one can write Arriaga's standard decomposition method like as per
eq\textasciitilde{}\eqref{eq_arriaga1}. To keep things simple, we assume
a lifetable radix of 1, meaning \(\ell_0 = 1\).

\[
\begin{aligned}
{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
 &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
\end{aligned}
\]

@@

\hypertarget{declarations}{%
\section*{Declarations}\label{declarations}}
\addcontentsline{toc}{section}{Declarations}

Some journals require declarations to be submitted in a standardised
format. Please check the Instructions for Authors of the journal to
which you are submitting to see if you need to complete this section. If
yes, your manuscript must contain the following sections under the
heading `Declarations':

\begin{itemize}
\tightlist
\item
  Funding
\item
  Conflict of interest/Competing interests (check journal-specific
  guidelines for which heading to use)
\item
  Ethics approval
\item
  Consent to participate
\item
  Consent for publication
\item
  Availability of data and materials
\item
  Code availability
\item
  Authors' contributions
\end{itemize}

\noindent If any of the sections are not relevant to your manuscript,
please include the heading and write `Not applicable' for that section.

\begin{flushleft}
Editorial Policies for:

\noindent Springer journals and proceedings:
\url{https://www.springer.com/gp/editorial-policies}

\noindent Nature Portfolio journals:
\url{https://www.nature.com/nature-research/editorial-policies}

\noindent \textit{Scientific Reports}:
\url{https://www.nature.com/srep/journal-policies/editorial-policies}

\noindent BMC journals:
\url{https://www.biomedcentral.com/getpublished/editorial-policies}

\end{flushleft}

\begin{appendices}

\hypertarget{secA1}{%
\section{Section title of first appendix}\label{secA1}}

@@

For submissions to Nature Portfolio Journals please use the heading
``Extended Data''.

\end{appendices}

\bibliography{bibliography.bib}

\end{document}
